GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis - sensitivity analysis

### 2023-10-12 figs/GSE158422_sensitivity.Rmd
maybe a competitive test is better?

boxplot(list("tumor"=matrix(colMeans(mval),ncol=2)[,1],"normal"=matrix(colMeans(mval),ncol=2)[,2]),
  main="mean probe methylation mval")

@@

@@

@@

Downsample

@@

```{r,gsamethdownsample}

@@

@@

@@

v1 <- list("AL up"=alsig_up,"AL dn"=alsig_dn,
  "GSAmeth up"=gsig_up,"GSAmeth dn"=gsig_dn)

>>>>

mval_means <- list("tumor"=matrix(colMeans(mval),ncol=2)[,1],"normal"=matrix(colMeans(mval),ncol=2)[,2])
boxplot(mval_means, main="mean probe methylation mval",cex=0,col="white")
beeswarm(mval_means,add=TRUE,pch=1,cex=1.5)
lapply(mval_means,mean)

@@

We will assume mean probe methylation is a proxy measure for global methylation.
The boxplot appears to show higher values in tumour samples, based on median and quartiles.
However the mean values are actually lower in tumour samples compared to normal.

@@

medf_means <- list("tumor"=matrix(colMeans(medf),ncol=2)[,1],"normal"=matrix(colMeans(medf),ncol=2)[,2])
boxplot(medf_means, main="mean gene methylation mval",cex=0,col="white")
beeswarm(medf_means,add=TRUE,pch=1,cex=1.5)

lapply(medf_means,mean)

@@

Try a competitive test.

```{r,compet}

mres <- mitch_calc(x=m,minsetsize=5,genesets=gs_symbols,cores=32,priority="effect")

# hypermethylated
mres_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)
nrow(mres_up)
head(mres_up,20) %>%
  kbl(caption="hypermethylated pathways with competitive test") %>%
  kable_paper("hover", full_width = F)

# hypomethylated
mres_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)
nrow(mres_dn)
head(mres_dn,20) %>%
  kbl(caption="hypomethylated pathways with competitive test") %>%
  kable_paper("hover", full_width = F)

These results look more in line with cancer pathogenesis.

Downsample AL test

@@

```{r,gsamethdownsample,fig.height=10}

@@

par(mfrow=c(1,1))

@@

msig_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)$set
msig_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)$set

@@

v1 <- list("AL up"=alsig_up,"AL dn"=alsig_dn,
  "mitch up"=msig_up,"mitch dn"=msig_dn,
  "GSAmeth up"=gsig_up,"GSAmeth dn"=gsig_dn)
